\newpage
\vspace*{5em}

\begin{center}
{\Huge {\bf Abstract}}
\end{center}
\vskip 4em

The Casimir effect constitutes one of the clearest and most direct manifestations of the quantum nature of the vacuum, emerging from the quantization of fields in the presence of boundary conditions that restrict the degrees of freedom of the system. Within the framework of quantum field theory, the vacuum is not a trivial state but is characterized by quantum fluctuations associated with the field modes, whose structure depends sensitively on the geometry and imposed constraints. In this work, we investigate the Casimir effect in the context of confined free scalar fields, emphasizing the fundamental role of boundary conditions in modifying the mode spectrum and, consequently, the vacuum energy.

We begin with a systematic review of the foundations of relativistic quantum field theory, focusing on the Lagrangian formulation and the derivation of equations of motion from the variational principle. The role of continuous symmetries and their connection to conserved quantities via Noether’s theorem is discussed, establishing the conceptual basis required for a rigorous definition of the energy and momentum of the field. This theoretical framework provides the necessary foundation for the canonical quantization of the scalar field, in which field modes are promoted to operators and the system’s energy is interpreted as a sum over independent harmonic oscillators.

Canonical quantization reveals the presence of zero-point energy associated with each field mode, which, in principle, leads to divergent sums. We show that, in the absence of boundary conditions, such divergences can be interpreted as unobservable background contributions. However, when the field is subject to geometric constraints, as in the case of parallel plates, the mode spectrum becomes discrete, and finite differences in vacuum energy acquire physical meaning. In this context, we develop the formalism required to handle these divergent sums, implicitly highlighting the need for regularization and subtraction procedures in order to extract physically meaningful quantities.

As a direct application of the developed formalism, we analyze the Casimir effect for a scalar field confined between parallel plates, imposing boundary conditions that lead to the quantization of modes along the direction perpendicular to the plates. The Casimir energy is computed as the difference between the vacuum energy of the confined system and that of free space, yielding expressions that explicitly depend on the separation between the plates. From this energy, we derive the Casimir pressure, demonstrating the existence of an effective force between the plates that arises purely from quantum vacuum fluctuations.

The obtained results clearly demonstrate the dependence of vacuum energy on geometry and boundary conditions, reinforcing the role of the Casimir effect as a fundamental conceptual tool for understanding the structure of the quantum vacuum. Furthermore, the formalism presented here provides a robust basis for generalizations, including the study of more complex geometries, the inclusion of external fields, and the consideration of different types of boundary conditions. Such extensions open perspectives for future investigations both in the theoretical domain—such as the study of spacetime topology effects and theories beyond the Standard Model—and in the experimental domain, where the Casimir effect plays a significant role in micro- and nanoscale systems.

In this way, this work contributes to consolidating the understanding of the Casimir effect within the framework of quantum field theory, highlighting its importance not only as a measurable physical phenomenon but also as a privileged window into the fundamental properties of the quantum vacuum.

\vspace*{2em}


\textbf{Keywords}: Casimir effect; vacuum energy; quantum field theory; scalar fields; boundary conditions.